---
title: "Visualizations"
format: html
---

## Setup

```{r}
#| label: setup
#| echo: false
#loading all required packages
library(tidyverse)
```

```{r}
#| label: reading-data
#| echo: false
#reading in all of the cleaned data
waste_population_q1_q2 <-read_csv("data-clean/waste-population-q1-q2")
waste_summary_q3 <- read_csv("data-clean/waste-summary-q3")
```

## Research Question 3

How has waste generation changed from 1995 in Monterey County?

```{r}
#| label: visualization-q3
#making a visualization to answer research question 3

waste_summary_q3 |>
ggplot(aes(x = time,
                y = disposal_ton, 
                color = city)) +
  geom_line() +
  labs(title = "Waste Generation in Monterey County by City (1995-2019)",
       color = "City",
       x = "Year",
       y = "Disposal Amount (tons)",
       alt = "The graph displays waste generation in Monterey County by city from 1995 to 2019. The line graph has year on the x axis and disposal amount in tons on the y axis, the lines are grouped by city. The lines are variable, often spiking in unision yearly. The overall trend shows a dip from around 2010 to 2015.")
```
The graph shows that waste generation has varied overtime with a high level of variability each year. It is hard to tell overall long term trends from this graph. There seems to be a dip from 2010 to 2015 however this could be due to the end of data on a number of cities and the start of data from the Salinas Valley Solid Waste Authority.
