---
title: "Visualizations"
format: html
---

## Setup

```{r}
#| label: setup
#| echo: false
#loading all required packages
library(tidyverse)
```

```{r}
#| label: reading-data
#| echo: false
#reading in all of the cleaned data
waste_population_q1_q2 <-read_csv("data-clean/waste-population-q1-q2")
waste_summary_q3 <- read_csv("data-clean/waste-summary-q3")
```

## Research Question 3

How has waste generation changed from 1995 in Monterey County?

```{r}
#| label: visualization-q3
#making a visualization to answer research question 3

waste_summary_q3 |>
ggplot(aes(x = time,
                y = disposal_ton, 
                color = fct_reorder2(city, time, disposal_ton))) +
  geom_line() +
  labs(title = "Waste Generation in Monterey County by City (1995-2019)",
       color = "City",
       x = "Year",
       y = "Disposal Amount per Quarter (Tons)",
       alt = "The graph displays waste generation in Monterey County by city from 1995 to 2019. The line graph has year on the x axis and disposal amount in tons on the y axis, the lines are grouped by city. The lines are variable, often spiking in unision yearly. The overall trend shows a dip from around 2010 to 2015. Data from CalRecycle.",
       caption = "Waste Disposal amounts in Monterey County each quarter from 1995 to 2019.
Data from CalRecycle") +
  scale_color_viridis_d()
```

The graph shows that waste generation has varied overtime with a high level of variability each year. There seems to be a dip from 2010 to 2015, however this could be due to the end of data on a number of cities.
```{r}
#| label: start-salinas-valley-solid-waste-authority
#figuring out exactly when Salinas Valley Solid Waste Authority reporting starts and cities end. For interpreting q3 graph and future graphs.
waste_summary_q3 |>
  filter(city == "Salinas Valley Solid Waste Authority") |>
  arrange(city)

waste_summary_q3 |>
filter(time == 2012.75)

waste_summary_q3 |>
filter(time == 2013)
```
In 2013 data for Gonzales, Greenfeild, King City, Salinas, and Soledad is no longer reported and data for the Salinas Valley Solid Waste Authority begins.

## Research Question 1

Question 1: How does population size of a city, for cities over 5000 people, affect waste generation per person in Monterey county?

```{r}
#| label: visualization-q1-1
waste_population_q1_q2 |>
  ggplot(aes(x = time, 
             y = disposal_per_population_ton)) +
  geom_point(aes(size = population_census_april_1_2010, 
                 color = fct_reorder2(city, time, disposal_per_population_ton))) +
  scale_color_viridis_d() +
  labs(color = "City",
       size = "City Population (2010)",
       title = " Waste Generation per Person in Monterey County (1995-2019)",
       caption = "Waste Generation per Person based on the 2010 census, for cities over 5000 people in Monterey 
       County by city and population size
       Data from US Census Bureau and CalRecycle",
       y = "Disposal Amount per Person per Quarter (Tons)",
       x = "Year",
       alt = "A graph displaying waste generation per person based on the 2010 census for cities over 5000 people in Monterey County, from 1995 to 2019. Waste disposal is measured in amount per person per quarter in tons. The graph shows Monterey, with a medium population size, gennerally has the highest disposal amount per person. Soledad and Greenfeild with similar polulation sizes generally had the lowest disposal amount per person. The largest city, Salinas has a medium waste generation amount per person. All cities experiance a dip in waste generation from roughly 2010 to 2015. Data for Gonzales, Greenfeild, King City, Salinas, and Soledad stops in 2013. Data from US Census Bureau and CalRecycle."
       )
```
The graph shows Salinas to be considerably larger in population than other cities. The graph shows a dip in waste generation per person between 2010 and 2015.

```{r}
#| label: visualization-q1-2
waste_population_q1_q2 |>
  ggplot(aes(x = population_census_april_1_2010,
             y = disposal_per_population_ton)) +
  geom_boxplot(aes(group = city, color = fct_reorder(city, desc(disposal_per_population_ton)))) +
  geom_smooth(method = lm, color = "black") +
  labs(color = "City")
```
The box plot shows a slight increase in  waste generation per person as population size increases.

## Research Question 2

How does population size per square mile affect waste generation per person for cities over 5000 people in Monterey County?

```{r}
#| label: visualization-q2-1
#chose to make population density into a color variable instead of a size variable because the size was causing too much overlapping. Chose not to break up by city because it was too hard to tell the different shapes.
waste_population_q1_q2 |>
  ggplot(aes(x = time, 
             y = disposal_per_population_ton)) +
  geom_point(aes(color = population_per_square_mile_2010)) +
  theme_light() +
  scale_color_viridis_c() +
  labs(title = "Waste per Person Based on Population Density Over Time in Monterey County",
       x = "Year",
       y = "Waste Disposal per Person per Quarter (Tons)",
       color = "Population Density 
per Square Mile (2010)",
      )
```

```{r}
#| label: visualization-q2-2
waste_population_q1_q2 |>
  ggplot(aes(x = population_per_square_mile_2010,
             y = disposal_per_population_ton)) +
  geom_boxplot(aes(group = city, 
             color = city)) +
  geom_smooth(color = "pink3", method = lm) +
  labs(title = "Population Density Versus Waste per Person by City",
       x = "Population Density per Square Mile (2010)",
       y = "Waste Disposal per Person per Quarter (Tons)",
       color = "City") +
  theme_light() +
  scale_color_viridis_d()
```
Because data for some cities stops and is grouped into The Salinas Valley Solid Waste Athority this could affect the graph.

```{r}
#| label: visualization-q2-3
#visualization to remove data after 2010 because that is when the Salinas Valley Solid Waste Authority data begins.
waste_population_q1_q2 |>
  filter(time < 2013) |>
  ggplot(aes(x = population_per_square_mile_2010,
             y = disposal_per_population_ton)) +
  geom_boxplot(aes(group = city, 
             color = city)) +
  geom_smooth(color = "pink3", method = lm) +
  labs(title = "Population Density Versus Waste per Person by City Prior to 2013",
       x = "Population Density per Square Mile (2010)",
       y = "Waste Disposal per Person per Quarter (Tons)",
       color = "City") +
  theme_light() +
  scale_color_viridis_d()
```