---
title: "Visualizations"
format: html
---

## Setup

```{r}
#| label: setup
#| echo: false
#loading all required packages
library(tidyverse)
```

```{r}
#| label: reading-data
#| echo: false
#reading in all of the cleaned data
waste_population_q1_q2 <-read_csv("data-clean/waste-population-q1-q2")
waste_summary_q3 <- read_csv("data-clean/waste-summary-q3")
```
## Research Question 1

```{r}
#| label: visualization-q1-1
waste_population_q1_q2 |>
  ggplot(aes(x = time, 
             y = disposal_per_population_ton)) +
  geom_point(aes(size = population_census_april_1_2010, 
                 color = city))
```

```{r}
#| label: visualization-q1-2
waste_population_q1_q2 |>
  ggplot(aes(x = population_census_april_1_2010,
             y = disposal_per_population_ton,
             group = city, 
             color = city)) +
  geom_boxplot()
```
```{r}
#| label: visualization-q2-1
#chose to make population density into a color variable instead of a size variable because the size was causing too much overlapping. Chose not to break up by city because it was too hard to tell the different shapes.
waste_population_q1_q2 |>
  ggplot(aes(x = time, 
             y = disposal_per_population_ton)) +
  geom_point(aes(color = population_per_square_mile_2010)) +
  theme_light() +
  scale_color_viridis_c() +
  labs(title = "Waste per Person Based on Population Density Over Time in Monterey County",
       x = "Year",
       y = "Waste Disposal per Person (Tons)",
       color = "Population Density 
per Square Mile (2010)",
      )
```

```{r}
#| label: visualization-q2-2
waste_population_q1_q2 |>
  ggplot(aes(x = population_per_square_mile_2010,
             y = disposal_per_population_ton)) +
  geom_boxplot(aes(group = city, 
             color = city)) +
  geom_smooth(color = "pink3") +
  labs(title = "Population Density Versus Waste per Person by City",
       x = "Population Density per Square Mile (2010)",
       y = "Waste Disposal per Person (Tons)",
       color = "City") +
  theme_light() +
  scale_color_viridis_d()
```

## Research Question 3

How has waste generation changed from 1995 in Monterey County?

```{r}
#| label: visualization-q3
#making a visualization to answer research question 3

waste_summary_q3 |>
ggplot(aes(x = time,
                y = disposal_ton, 
                color = city)) +
  geom_line() +
  labs(title = "Waste Generation in Monterey County by City (1995-2019)",
       color = "City",
       x = "Year",
       y = "Disposal Amount (tons)",
       alt = "The graph displays waste generation in Monterey County by city from 1995 to 2019. The line graph has year on the x axis and disposal amount in tons on the y axis, the lines are grouped by city. The lines are variable, often spiking in unision yearly. The overall trend shows a dip from around 2010 to 2015.")
```
The graph shows that waste generation has varied overtime with a high level of variability each year. It is hard to tell overall long term trends from this graph. There seems to be a dip from 2010 to 2015 however this could be due to the end of data on a number of cities and the start of data from the Salinas Valley Solid Waste Authority.
